---
title: "Day 4 Lecture"
output: html_document
---

# Today's agenda

  - Review & questions
  - Character/string data
  - Factors
  - Excursion: simple data frame manipulation, `xtabs()`
  - PRACTICE: working with factors, part 1
  - Special values: NA/NULL/Inf/NaN
  - PRACTICE: data type coersion
  - Lists
  - Creating data frames from scratch
  - Combining data frames
  - PRACTICE: binding & merging

# Review
  - Go over Day 3 Practice
  - Day 3 Homework example
  - Knitting notes

# Characters
  - more primitive than factors
  - can manipulate with regular expressions, other text functions

Some handy text stuff

```{r}
# quotes
x <- "a string" #scott tends to use double quotes bc r uses double quotes when it shows you what it did in the console
x
y <- 'also a string'
y
z <- "this is a 'string' also"
z

# cat(), print() 
print(x) #shows x as a vector with [1] and quotes... gets you what you need but doesn't look as clean as cat(x)
cat(x) #just prints string without quotes so it looks like normal text


#print is acutally always happening behind the scenes when you just type
2+2 #same as print(2+2)
cat(2+2) #would get rid of [1] in console

# special characters: \t, \n, "escaping"

x <- "here's a cool result: 10"
cat(x)

y <- "here's a cool result:\t10"
cat(y)

z <- "here's a cool result:\n\n\t\t\t10"
cat(z)

#\t for tab
#\n for line returns
cat("some\ntext")
cat("some\\ntext") #extra \ tells R to treat \ as \ rather than sepcial character


# nchar() (#length refers to a vector so if you say length of hello it's return 1 if you want the length of a string use nchar and it'd return 5)
nchar("hello")
length("hello")

# paste(), strsplit()
paste ("hello", "there")
?paste #paste defaults to a space separator
paste ("hello", "there", sep = "") #this would get rid of the space sperator

x<- c("here", "are", "some", "strings")
x

paste("prefix", x) #this would put "prefix" in front of every instance of x

paste(x, collapse = " ") #collapses everything into one string separated by a space

x
y <- paste (x, "suffix", sep = "_")
y

strsplit(y, split = "_") #good to use when there's an easy to see delimiter like - but if there isn't an easy delimiter use gsub and figure out the pattern

z <- paste(x, "suffix", sep = ".")
z

strsplit(z, split = ".") #WONT WORK BC . MEANS EVERY CHACARACTER
strsplit(z, split = "\\.") #use the \\ to tell R you want to use the . as a delimiter


# stringr package #use this if you do a lot of data manipulation with text / collapsing

# grep(), grepl(), gsub()
x
grep("here", x) #here is pattern and x is vector and grep returns where in the string 
grep("e", x)


grepl("here", x) #grepl is the logical version of grep and will return TRUE or FALSE

###basic gsub
gsub("e", "E", x) #search and replace runction, give what you're looking for what to replace and the vector to look in
gsub ("e", "", x)

###fancy gsub
longer.thing <- c("10F5657", "59J4957", "30K16238")
gsub("\\d+\\w(\\d+)", "\\1", longer.thing) #d means digit and #w means word; this will just get digits
gsub("(\\d+)(\\w)(\\d+)", "\\2", longer.thing) #this will jsut get letters (the \\2 refers to the second set of parehtneticals so it'll replace the whole thing with what's in the second set of parehtniesis which is a w aka letters)

longer.thing2 <- c("10F5657ggjggjjg", "59J4957sfs", "30K16238eeeee")
gsub("(\\d+)(\\w)(\\d+)", "\\2", longer.thing2)


```

# Factors
  - Actually numbers under the hood
  - Act like numbers + labels
  - VERY useful for many kinds of analyses
  - if a column is a *property* of data (not a measure itself), factors are usually best
    - groupings = good as factors
    - conditions = good as factors
  - if a column is *data*, maybe consider keeping it a string
    - free response answers = bad as factors
  - `as.factor()` coerces to a factor
  - `factor()` creates a factor
  - factors have *levels*: categories
  - `stringsAsFactors` : option for how to read in character vectors
    - = FALSE if you don't want to read in characters as factors

```{r}
# factor() #this is the way to make a factor
x <- c(rep("a", 20), rep("b", 20), rep("c", 20)) #making a vector with A 20 times, B 20 time, c 20 times
x
summary(x) #shows list

x.factor <- factor(x)
summary (x.factor) #shows count of each A B and C


# levels()
levels(x.factor)


# summary()

# coersion

# subsetting
sleep
mysleep <- sleep
summary(mysleep) #extra is numbers group is factor you can tell based on the output

class(mysleep$group) #tells you what kind of data it is

levels(mysleep$group)

mysleep.group1 <- (mysleep [mysleep$group=="1",])
summary(mysleep.group1)

# droplevels
mysleep.group1 <- droplevels(mysleep.group1) #drop levels gets rid of factors that have no observations (in this case, group 2)
mysleep.group1


# changing levels (labels)
mysleep <- sleep
levels(mysleep$group)
levels(mysleep$group) <- c("pre", "post")
levels(mysleep$group)
summary(mysleep)
head(mysleep)

colnames(mysleep)
colnames(mysleep) <- c("extra sleep", "treatment group", "subject")
head(mysleep)


# re-ordering
mysleep <- sleep
levels(mysleep$group) <- c("pre", "post")
summary(mysleep)
#levels(mysleep$group) <- c("post", "post") #WRong. CHANGES DATA, NOT ORDER OF LEVELS
#levels(mysleep)

mysleep$group2 <- mysleep$group
head(mysleep)
mysleep$group2 <- factor(mysleep$group2, levels = c("post", "pre"))
summary(mysleep)
head(mysleep)


```

# Data manipulation excursion
## Creating (simple) new columns

```{r}
# single value
mysleep$experiment <- 1
summary(mysleep)
mysleep$group3 <- mysleep$group #make a new column by assigning values to that column
summary(mysleep)

#ifelse()
mysleep$sleeplevel <- ifelse(mysleep$extra >= 1, "high", "low") #straight if else statement
summary(mysleep)

mysleep$sleeplevel <- ifelse(mysleep$extra >= 1, "high", 
                             ifelse(mysleep$extra < 1, "low", NA)) #nested if else statement, adding NA is helpful for complicated stateements so it'll stand out to you if something in yoru code isn't working
summary(mysleep)

mysleep$sleeplevel <- as.factor(ifelse(mysleep$extra >= 1, "high", "low"))
summary(mysleep)

mysleep$sleeplevel <- factor(ifelse(mysleep$extra >= 1, "high", "low"), levels = c("low", "high"))
summary(mysleep)

# using subsetting

```

## Using xtabs() to check coding/cross-tabs

```{r}
#for crosstabs, helpful for data checks after you've manipulated data
xtabs(~ group + group2, data = mysleep) # ~ signals to R that you're doing a formula
xtabs(~ sleeplevel + extra, data = mysleep)
xtabs(~ extra + sleeplevel, data = mysleep) #same as above but different orientatino
```

# PRACTICE: #0 -- #6

# NAs, NULLs

## NAs
  - missing data... means unknown; if age is missing the person has an age you just don't know it so you have to decide if/how to impute
  - can be a huge pain
  - hugely important to handle properly!
  - no universal "handle NAs" or "ignore NAs" argument
  - sometimes `na.rm = TRUE` (like for `mean()`)
  - sometimes it's other stuff
  - `cor()` use = "pairwise.complete.obs"
  - no universal "deal with NAs" argument
  
```{r}

```

## NULL
  - another special value
  - means "nothing" or "doesn't exist" (contrast with "missing" or "unknown")

```{r}

```

## Inf, -Inf, NaN
  - can sometimes result from mathematical operations
  - helpful to be able to identify, locate them

```{r}
x <- 1/0
x

y <- -1/0
y

z <- sqrt(-4)
z


```

# Data types, classes, & coersion

```{r}
# typeof()

# class()

```

`is.X` family of functions (`is.factor()`, `is.na()`, `is.finite()`, etc.)
`as.X` family of functions (`as.factor()`, `as.numeric()`, `as.data.frame()`, etc.)

# PRACTICE: #7

# Lists
  - similar to vectors
  - can contain different data types
  - use [[ ]] to pull out values
  - handy for glomming large/complex objects together
  - data frames are special lists!
  
```{r}

```

# Creating data frames
  - handy, especially for making "look-up tables"
  
```{r}
# data.frame()

```

# PRACTICE: #8 -- #11

# Combining data frame objects
## `rbind()` and `cbind()`
  - rbind
    - glues rows on top of each other
    - expects same columns
    - new and improved `bind_rows()` in `dplyr` package
  - cbind
    - glues columns next to each other
    - need to be the same length    
```{r}

```

## `merge()` and `*_join()` from `dplyr` package
```{r}

```
    - simplest: have ONE column in common, named the same thing (with the same values)
    - can get MUCH more complex
    - `nrow()` good check on what was gained/lost during a merge
    - "join" functions in `dplyr`
      - `inner_join()` is like all = FALSE
      - `full_join()` is like all = TRUE
      - `left_join()`, `right_join()`, etc.
      - see ?join for more details



# PRACTICE: #12 -- #14


